---
title: "Metagenome Annotation Pipeline — QC Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  ann_dir:   NULL
  metadata:  "/Users/peterfalb/Documents/Research/Kennedy/ForestGEO/MesoFun/stakes.csv"
  min_tools: 2
  out_dir:   NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(kableExtra)
})

if (is.null(params$ann_dir)) stop("params$ann_dir is required.")

out_dir <- if (is.null(params$out_dir)) file.path(params$ann_dir, "qc") else params$out_dir
fig_dir <- file.path(out_dir, "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

theme_set(theme_bw(base_size = 11) + theme(
  axis.text.x = element_text(angle = 45, hjust = 1),
  strip.background = element_rect(fill = "#f0f0f0")
))

save_fig <- function(p, name, width = 12, height = 5) {
  ggsave(file.path(fig_dir, paste0(name, ".png")), plot = p,
         width = width, height = height, dpi = 150)
  p
}

# Palette for stand_type
stand_colors <- c(
  "ash or elm" = "#66C2A5",
  "oak"        = "#FC8D62",
  "pine"       = "#8DA0CB",
  "mixed"      = "#E78AC3",
  "unknown"    = "#CCCCCC"
)
```

```{r load-samples}
sample_list_file <- file.path(params$ann_dir, "sample_list.txt")
if (!file.exists(sample_list_file)) {
  stop("sample_list.txt not found: ", sample_list_file)
}
samples <- readLines(sample_list_file)
samples <- samples[nzchar(trimws(samples))]
```

```{r load-metadata}
meta_df <- NULL
sample_meta <- data.frame(
  sample     = samples,
  stake_id   = sub("^GEO-(.+)_S[0-9]+$", "\\1", samples),
  stand_type = "unknown",
  stringsAsFactors = FALSE
)

if (!is.null(params$metadata) && file.exists(params$metadata)) {
  meta_raw <- read.csv(params$metadata, stringsAsFactors = FALSE)
  meta_df  <- meta_raw[meta_raw$metagenomics == "yes", c("stake", "stand_type")]
  sample_meta <- sample_meta %>%
    left_join(meta_df, by = c("stake_id" = "stake")) %>%
    mutate(stand_type = coalesce(stand_type.y, "unknown")) %>%
    select(sample, stake_id, stand_type)
}

add_stand_facet <- function() {
  if (!is.null(meta_df)) facet_wrap(~stand_type, scales = "free_x", nrow = 1) else NULL
}
```

```{r define-loaders}
# ── Tiara summary ─────────────────────────────────────────────────────────────
load_tiara <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "tiara", paste0(s, "_tiara_summary.txt"))
    if (!file.exists(f) || file.size(f) == 0) {
      return(data.frame(sample = s, n_total = NA_integer_, n_euk = NA_integer_,
                        n_prok = NA_integer_, n_mito = NA_integer_,
                        n_plastid = NA_integer_, n_unknown = NA_integer_,
                        stringsAsFactors = FALSE))
    }
    lines <- readLines(f)
    parse_val <- function(pattern) {
      line <- grep(pattern, lines, value = TRUE, ignore.case = TRUE)
      if (length(line) == 0) return(NA_integer_)
      m <- regexpr("[0-9]+", line[1])
      if (m == -1) NA_integer_ else as.integer(regmatches(line[1], m))
    }
    data.frame(
      sample    = s,
      n_total   = parse_val("Total contigs"),
      n_euk     = parse_val("Eukarya"),
      n_prok    = parse_val("Prokarya"),
      n_mito    = parse_val("Mitochondria"),
      n_plastid = parse_val("Plastid"),
      n_unknown = parse_val("Unknown"),
      stringsAsFactors = FALSE
    )
  }) %>% bind_rows()
}

# ── MetaEuk protein count ─────────────────────────────────────────────────────
load_metaeuk <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "metaeuk", s, paste0(s, ".fas"))
    n <- if (file.exists(f) && file.size(f) > 0) {
      as.integer(trimws(system(
        paste0("grep -c '>' ", shQuote(f), " 2>/dev/null || echo 0"),
        intern = TRUE
      )[1]))
    } else NA_integer_
    data.frame(sample = s, n_proteins = n, stringsAsFactors = FALSE)
  }) %>% bind_rows()
}

# ── KOfamScan ─────────────────────────────────────────────────────────────────
load_kofam <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "kofam", paste0(s, "_kofam_mapper.tsv"))
    if (!file.exists(f) || file.size(f) == 0) {
      return(data.frame(sample = s, n_ko_genes = NA_integer_, stringsAsFactors = FALSE))
    }
    dt <- tryCatch(
      fread(f, sep = "\t", header = FALSE, comment.char = "#",
            col.names = c("ko", "gene_id")),
      error = function(e) NULL
    )
    if (is.null(dt) || nrow(dt) == 0) {
      return(data.frame(sample = s, n_ko_genes = 0L, stringsAsFactors = FALSE))
    }
    data.frame(sample = s, n_ko_genes = length(unique(dt$gene_id)),
               stringsAsFactors = FALSE)
  }) %>% bind_rows()
}

# ── dbCAN ─────────────────────────────────────────────────────────────────────
find_dbcan_overview <- function(dir) {
  for (nm in c("overview.tsv", "overview.txt", "CAZyme_annotation.tsv")) {
    f <- file.path(dir, nm)
    if (file.exists(f) && file.size(f) > 0) return(f)
  }
  NULL
}

load_dbcan <- function(samples, ann_dir, min_tools = 2) {
  lapply(samples, function(s) {
    f <- find_dbcan_overview(file.path(ann_dir, "dbcan", s))
    if (is.null(f)) {
      return(data.frame(sample = s, cazy_class = NA_character_,
                        n_hicof = NA_integer_, stringsAsFactors = FALSE))
    }
    dt <- tryCatch(fread(f, sep = "\t"), error = function(e) NULL)
    if (is.null(dt) || nrow(dt) == 0) {
      return(data.frame(sample = s, cazy_class = "none",
                        n_hicof = 0L, stringsAsFactors = FALSE))
    }

    hmm_col   <- grep("hmm",              names(dt), value = TRUE, ignore.case = TRUE)[1]
    tools_col <- grep("#.*tool|ofTool|n_tool", names(dt), value = TRUE, ignore.case = TRUE)[1]

    if (is.na(hmm_col) || is.na(tools_col)) {
      return(data.frame(sample = s, cazy_class = NA_character_,
                        n_hicof = NA_integer_, stringsAsFactors = FALSE))
    }

    dt_hi <- dt[dt[[tools_col]] >= min_tools, ]
    if (nrow(dt_hi) == 0) {
      return(data.frame(sample = s, cazy_class = "none",
                        n_hicof = 0L, stringsAsFactors = FALSE))
    }

    hmm_vals <- as.character(dt_hi[[hmm_col]])
    hmm_vals[hmm_vals %in% c("N/A", "-", "", "NA")] <- NA_character_
    class_vals <- sub("([A-Z]+).*", "\\1", hmm_vals)
    class_vals[!class_vals %in% c("GH", "AA", "CE", "PL", "CBM", "GT")] <- "other"
    class_vals[is.na(hmm_vals)] <- "unclassified"

    result <- as.data.frame(table(class_vals), stringsAsFactors = FALSE)
    names(result) <- c("cazy_class", "n_hicof")
    result$sample <- s
    result
  }) %>% bind_rows()
}

# ── PHI-base ──────────────────────────────────────────────────────────────────
load_phibase <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "phibase", paste0(s, "_phibase.tsv"))
    if (!file.exists(f) || file.size(f) == 0) {
      return(data.frame(sample = s, phi_phenotype = NA_character_,
                        n_hits = NA_integer_, stringsAsFactors = FALSE))
    }
    dt <- tryCatch(fread(f, sep = "\t", header = FALSE), error = function(e) NULL)
    if (is.null(dt) || nrow(dt) == 0) {
      return(data.frame(sample = s, phi_phenotype = "none",
                        n_hits = 0L, stringsAsFactors = FALSE))
    }

    # Find column containing '#'-delimited stitle (PHI-base header format)
    stitle_col <- which(sapply(dt, function(x) any(grepl("#", head(x, 10)))))[1]
    if (is.na(stitle_col)) {
      return(data.frame(sample = s, phi_phenotype = "unknown",
                        n_hits = nrow(dt), stringsAsFactors = FALSE))
    }

    phenotypes <- sapply(as.character(dt[[stitle_col]]), function(x) {
      parts <- strsplit(x, "#")[[1]]
      if (length(parts) >= 6) trimws(parts[6]) else "unknown"
    })

    phenotypes <- case_when(
      str_detect(tolower(phenotypes), "reduced.{0,10}virulen")    ~ "reduced_virulence",
      str_detect(tolower(phenotypes), "loss.{0,10}pathogen")      ~ "loss_of_pathogenicity",
      str_detect(tolower(phenotypes), "increased.{0,10}virulen")  ~ "increased_virulence",
      str_detect(tolower(phenotypes), "effector")                 ~ "effector",
      str_detect(tolower(phenotypes), "unaffected")               ~ "unaffected",
      TRUE                                                        ~ "other"
    )

    result <- as.data.frame(table(phenotypes), stringsAsFactors = FALSE)
    names(result) <- c("phi_phenotype", "n_hits")
    result$sample <- s
    result
  }) %>% bind_rows()
}

# ── MMseqs2 LCA ───────────────────────────────────────────────────────────────
load_mmseqs <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "mmseqs_taxonomy", paste0(s, "_lca.tsv"))
    if (!file.exists(f) || file.size(f) == 0) {
      return(data.frame(sample = s, kingdom_label = NA_character_,
                        n = NA_integer_, stringsAsFactors = FALSE))
    }
    dt <- tryCatch(
      fread(f, sep = "\t", header = FALSE,
            col.names = c("gene_id", "taxid", "rank", "name", "lineage")),
      error = function(e) NULL
    )
    if (is.null(dt) || nrow(dt) == 0) {
      return(data.frame(sample = s, kingdom_label = "unclassified",
                        n = 0L, stringsAsFactors = FALSE))
    }

    dt[, kingdom_label := case_when(
      taxid == 0 | is.na(taxid)              ~ "unclassified",
      str_detect(lineage, "Ascomycota")      ~ "Ascomycota",
      str_detect(lineage, "Basidiomycota")   ~ "Basidiomycota",
      str_detect(lineage, "Glomeromycota")   ~ "Glomeromycota",
      str_detect(lineage, "Fungi")           ~ "other Fungi",
      str_detect(lineage, "Eukaryota")       ~ "other Eukaryota",
      str_detect(lineage, "Bacteria")        ~ "Bacteria",
      str_detect(lineage, "Archaea")         ~ "Archaea",
      TRUE                                   ~ "unclassified"
    )]

    result <- dt[, .N, by = kingdom_label]
    result[, sample := s]
    as.data.frame(result)
  }) %>% bind_rows()
}

# ── featureCounts summary ─────────────────────────────────────────────────────
load_featurecounts <- function(samples, ann_dir) {
  lapply(samples, function(s) {
    f <- file.path(ann_dir, "featurecounts", paste0(s, "_counts.txt.summary"))
    if (!file.exists(f) || file.size(f) == 0) {
      return(data.frame(sample = s, status = NA_character_,
                        count = NA_integer_, stringsAsFactors = FALSE))
    }
    dt <- tryCatch(fread(f, sep = "\t", header = TRUE), error = function(e) NULL)
    if (is.null(dt) || nrow(dt) == 0) {
      return(data.frame(sample = s, status = NA_character_,
                        count = NA_integer_, stringsAsFactors = FALSE))
    }
    setnames(dt, c("status", "count"))
    dt[, sample := s]
    as.data.frame(dt)
  }) %>% bind_rows()
}
```

```{r run-loaders, cache=FALSE}
tiara_df   <- load_tiara(samples, params$ann_dir)
metaeuk_df <- load_metaeuk(samples, params$ann_dir)
kofam_df   <- load_kofam(samples, params$ann_dir)
dbcan_df   <- load_dbcan(samples, params$ann_dir, min_tools = params$min_tools)
phi_df     <- load_phibase(samples, params$ann_dir)
mmseqs_df  <- load_mmseqs(samples, params$ann_dir)
fc_df      <- load_featurecounts(samples, params$ann_dir)

# Attach stand_type to base dataframes
tiara_df   <- left_join(tiara_df,   sample_meta, by = "sample")
metaeuk_df <- left_join(metaeuk_df, sample_meta, by = "sample")
```

---

## Data availability

```{r missing-data-check}
check_missing <- function(df, step) {
  sub_df <- df[, !names(df) %in% c("sample", "stand_type", "stake_id"), drop = FALSE]
  missing <- df$sample[apply(sub_df, 1, function(r) all(is.na(r)))]
  data.frame(step = step, n_missing = length(missing),
             samples = if (length(missing) > 0) paste(missing, collapse = ", ") else "—",
             stringsAsFactors = FALSE)
}

avail <- bind_rows(
  check_missing(tiara_df,                                                     "Tiara"),
  check_missing(metaeuk_df,                                                   "MetaEuk"),
  check_missing(kofam_df,                                                     "KOfamScan"),
  check_missing(dbcan_df %>% group_by(sample) %>%
                  summarise(n_hicof = if (all(is.na(n_hicof))) NA_integer_ else sum(n_hicof, na.rm=TRUE),
                            .groups="drop"), "dbCAN"),
  check_missing(phi_df %>% group_by(sample) %>%
                  summarise(n_hits = if (all(is.na(n_hits))) NA_integer_ else sum(n_hits, na.rm=TRUE),
                            .groups="drop"),  "PHI-base"),
  check_missing(mmseqs_df %>% group_by(sample) %>%
                  summarise(n = if (all(is.na(.data$n))) NA_integer_ else sum(.data$n, na.rm=TRUE),
                            .groups="drop"),  "MMseqs2 LCA"),
  check_missing(fc_df %>% group_by(sample) %>%
                  summarise(count = if (all(is.na(count))) NA_integer_ else sum(count, na.rm=TRUE),
                            .groups="drop"),  "featureCounts")
)

kable(avail, col.names = c("Step", "Missing samples", "Sample IDs"),
      caption = paste0("Samples with no output detected (n = ", length(samples), " total)")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  row_spec(which(avail$n_missing > 0), background = "#FFF3CD")
```

---

## Fig 1: Tiara Contig Classification {.tabset}

```{r fig1, fig.height=5}
tiara_long <- tiara_df %>%
  filter(!is.na(n_total)) %>%
  mutate(
    pct_euk     = n_euk     / n_total * 100,
    pct_prok    = n_prok    / n_total * 100,
    pct_mito    = n_mito    / n_total * 100,
    pct_plastid = n_plastid / n_total * 100,
    pct_unknown = n_unknown / n_total * 100
  ) %>%
  pivot_longer(starts_with("pct_"), names_to = "class", values_to = "pct") %>%
  mutate(class = factor(recode(class,
    pct_euk     = "Eukarya",
    pct_prok    = "Prokarya",
    pct_mito    = "Mitochondria",
    pct_plastid = "Plastid",
    pct_unknown = "Unknown"
  ), levels = c("Eukarya", "Prokarya", "Mitochondria", "Plastid", "Unknown")))

tiara_totals <- tiara_df %>%
  filter(!is.na(n_total)) %>%
  mutate(label = paste0(round(n_total / 1000, 1), "k"))

p1 <- ggplot(tiara_long, aes(x = sample, y = pct, fill = class)) +
  geom_col(position = "stack", width = 0.8) +
  geom_text(data = tiara_totals,
            aes(x = sample, y = 101, label = label, fill = NULL),
            size = 2.5, vjust = 0, inherit.aes = FALSE) +
  scale_fill_manual(values = c(
    "Eukarya"     = "#4DAF4A",
    "Prokarya"    = "#377EB8",
    "Mitochondria"= "#FF7F00",
    "Plastid"     = "#984EA3",
    "Unknown"     = "#AAAAAA"
  )) +
  add_stand_facet() +
  labs(title = "Tiara Contig Classification",
       subtitle = "Text labels = total contigs (thousands)",
       x = NULL, y = "% of contigs", fill = "Class") +
  coord_cartesian(ylim = c(0, 110))

save_fig(p1, "fig1_tiara")
```

---

## Fig 2: MetaEuk Gene Prediction {.tabset}

```{r fig2, fig.height=5}
med_p <- median(metaeuk_df$n_proteins, na.rm = TRUE)
q25_p <- quantile(metaeuk_df$n_proteins, 0.25, na.rm = TRUE)

p2 <- ggplot(metaeuk_df %>% filter(!is.na(n_proteins)),
             aes(x = sample, y = n_proteins, fill = stand_type)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = med_p, linetype = "dashed", linewidth = 0.6, color = "black") +
  geom_hline(yintercept = q25_p, linetype = "dotted", linewidth = 0.6, color = "#D7191C") +
  scale_fill_manual(values = stand_colors, na.value = "#CCCCCC") +
  add_stand_facet() +
  labs(title    = "MetaEuk Predicted Proteins per Sample",
       subtitle = paste0("Dashed = median (", round(med_p),
                         "); red dotted = 25th percentile (", round(q25_p), ")"),
       x = NULL, y = "# predicted proteins", fill = "Stand type")

save_fig(p2, "fig2_metaeuk")
```

---

## Fig 3: Annotation Coverage Heatmap {.tabset}

```{r fig3, fig.height=6, fig.width=9}
# Per-sample totals needed for % calculation
prot_total <- metaeuk_df %>% select(sample, n_proteins)

kofam_sum <- kofam_df %>% select(sample, n_ko_genes)

dbcan_sum <- dbcan_df %>%
  filter(!is.na(n_hicof)) %>%
  group_by(sample) %>%
  summarise(n_cazy = sum(n_hicof), .groups = "drop")

phi_sum <- phi_df %>%
  filter(!is.na(n_hits)) %>%
  group_by(sample) %>%
  summarise(n_phi = sum(n_hits), .groups = "drop")

mmseqs_sum <- mmseqs_df %>%
  filter(!is.na(.data$n)) %>%
  group_by(sample) %>%
  summarise(
    n_classified = sum(.data$n[kingdom_label != "unclassified"], na.rm = TRUE),
    n_fungi      = sum(.data$n[kingdom_label %in% c("Ascomycota", "Basidiomycota",
                                                     "Glomeromycota", "other Fungi")], na.rm = TRUE),
    .groups = "drop"
  )

heatmap_df <- prot_total %>%
  left_join(kofam_sum,  by = "sample") %>%
  left_join(dbcan_sum,  by = "sample") %>%
  left_join(phi_sum,    by = "sample") %>%
  left_join(mmseqs_sum, by = "sample") %>%
  left_join(sample_meta %>% select(sample, stand_type), by = "sample") %>%
  mutate(
    pct_kofam    = n_ko_genes   / n_proteins * 100,
    pct_cazy     = n_cazy       / n_proteins * 100,
    pct_phi      = n_phi        / n_proteins * 100,
    pct_mmseqs   = n_classified / n_proteins * 100,
    pct_fungi    = n_fungi      / n_proteins * 100
  )

# Sort samples by stand_type then by name
heatmap_df <- heatmap_df %>%
  arrange(stand_type, sample) %>%
  mutate(sample = factor(sample, levels = sample))

heatmap_long <- heatmap_df %>%
  select(sample, pct_kofam, pct_cazy, pct_phi, pct_mmseqs, pct_fungi) %>%
  pivot_longer(-sample, names_to = "tool", values_to = "pct") %>%
  mutate(tool = recode(tool,
    pct_kofam  = "KOfamScan",
    pct_cazy   = paste0("dbCAN\n(≥", params$min_tools, " tools)"),
    pct_phi    = "PHI-base",
    pct_mmseqs = "MMseqs2\nclassified",
    pct_fungi  = "MMseqs2\nFungi"
  ))

p3 <- ggplot(heatmap_long, aes(x = tool, y = sample, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(is.na(pct), "—", sprintf("%.1f%%", pct))),
            size = 2.8, color = "black") +
  scale_fill_gradient(low = "#EFF3FF", high = "#2171B5",
                      na.value = "gray92", name = "% genes") +
  labs(title    = "Annotation Coverage Heatmap",
       subtitle = "Samples sorted by stand type; fill = % of predicted proteins annotated",
       x = NULL, y = NULL)

save_fig(p3, "fig3_annotation_heatmap", width = 9, height = max(6, length(samples) * 0.25 + 2))
```

---

## Fig 4: CAZyme Class Breakdown {.tabset}

```{r fig4, fig.height=5}
cazy_colors <- c(
  "GH"           = "#E41A1C",
  "AA"           = "#377EB8",
  "CE"           = "#4DAF4A",
  "PL"           = "#FF7F00",
  "CBM"          = "#984EA3",
  "GT"           = "#A65628",
  "other"        = "#AAAAAA",
  "unclassified" = "#DDDDDD",
  "none"         = "#F0F0F0"
)

dbcan_plot <- dbcan_df %>%
  filter(!is.na(cazy_class), cazy_class != "none") %>%
  left_join(sample_meta %>% select(sample, stand_type), by = "sample")

total_range <- dbcan_df %>%
  filter(!is.na(n_hicof)) %>%
  group_by(sample) %>%
  summarise(total = sum(n_hicof)) %>%
  summarise(rng = paste0(min(total), "–", max(total)))

p4 <- ggplot(dbcan_plot, aes(x = sample, y = n_hicof,
             fill = factor(cazy_class, levels = names(cazy_colors)))) +
  geom_col(position = "stack", width = 0.8) +
  scale_fill_manual(values = cazy_colors, drop = FALSE) +
  add_stand_facet() +
  labs(title    = paste0("CAZyme Class Breakdown (≥", params$min_tools, " tools)"),
       subtitle = paste0("Total high-confidence CAZymes per sample: ", total_range$rng),
       x = NULL, y = "# high-confidence CAZyme calls", fill = "CAZy class")

save_fig(p4, "fig4_cazyme_classes")
```

---

## Fig 5: PHI-base Phenotype Breakdown {.tabset}

```{r fig5, fig.height=5}
phi_colors <- c(
  "reduced_virulence"     = "#FC8D62",
  "loss_of_pathogenicity" = "#E31A1C",
  "increased_virulence"   = "#FF7F00",
  "effector"              = "#6A3D9A",
  "unaffected"            = "#33A02C",
  "other"                 = "#AAAAAA"
)

phi_plot <- phi_df %>%
  filter(!is.na(phi_phenotype), phi_phenotype != "none") %>%
  left_join(sample_meta %>% select(sample, stand_type), by = "sample") %>%
  mutate(phi_phenotype = factor(phi_phenotype, levels = names(phi_colors)))

p5 <- ggplot(phi_plot, aes(x = sample, y = n_hits, fill = phi_phenotype)) +
  geom_col(position = "stack", width = 0.8) +
  scale_fill_manual(values = phi_colors, drop = FALSE) +
  add_stand_facet() +
  labs(title = "PHI-base Hit Phenotypes",
       x = NULL, y = "# DIAMOND hits", fill = "Phenotype")

save_fig(p5, "fig5_phibase_phenotypes")
```

---

## Fig 6: MMseqs2 Taxonomy {.tabset}

```{r fig6, fig.height=5}
kingdom_levels <- c("Ascomycota", "Basidiomycota", "Glomeromycota",
                    "other Fungi", "other Eukaryota",
                    "Bacteria", "Archaea", "unclassified")

kingdom_colors <- c(
  "Ascomycota"      = "#E8736C",
  "Basidiomycota"   = "#4DBEAE",
  "Glomeromycota"   = "#EDB120",
  "other Fungi"     = "#77AC30",
  "other Eukaryota" = "#7E2F8E",
  "Bacteria"        = "#AAAAAA",
  "Archaea"         = "#D4B896",
  "unclassified"    = "#DDDDDD"
)

mmseqs_plot <- mmseqs_df %>%
  filter(!is.na(kingdom_label)) %>%
  left_join(sample_meta %>% select(sample, stand_type), by = "sample") %>%
  group_by(sample, stand_type, kingdom_label) %>%
  summarise(n = sum(.data$n, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(kingdom_label = factor(kingdom_label, levels = kingdom_levels))

p6 <- ggplot(mmseqs_plot,
             aes(x = sample, y = pct, fill = kingdom_label)) +
  geom_col(position = "stack", width = 0.8) +
  scale_fill_manual(values = kingdom_colors, drop = FALSE) +
  add_stand_facet() +
  labs(title   = "MMseqs2 Protein-level Taxonomy (LCA)",
       caption = paste0("Bacterial assignments on eukaryotic contigs reflect conserved ",
                        "proteins, not contamination (all contigs pre-filtered by Tiara)."),
       x = NULL, y = "% of proteins", fill = "Kingdom / Phylum")

save_fig(p6, "fig6_mmseqs_taxonomy")
```

---

## Fig 7: featureCounts Read Assignment {.tabset}

```{r fig7, fig.height=5}
fc_colors <- c(
  "Assigned"                 = "#1F78B4",
  "Unassigned_NoFeatures"    = "#A6CEE3",
  "Unassigned_Ambiguity"     = "#FB9A99",
  "other"                    = "#DDDDDD"
)

fc_plot <- fc_df %>%
  filter(!is.na(status)) %>%
  mutate(status_grouped = case_when(
    status == "Assigned"              ~ "Assigned",
    status == "Unassigned_NoFeatures" ~ "Unassigned_NoFeatures",
    status == "Unassigned_Ambiguity"  ~ "Unassigned_Ambiguity",
    TRUE                              ~ "other"
  )) %>%
  group_by(sample, status_grouped) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(pct = count / sum(count) * 100) %>%
  ungroup() %>%
  left_join(sample_meta %>% select(sample, stand_type), by = "sample") %>%
  mutate(status_grouped = factor(status_grouped, levels = names(fc_colors)))

assigned_labels <- fc_plot %>% filter(status_grouped == "Assigned")

p7 <- ggplot(fc_plot, aes(x = sample, y = count, fill = status_grouped)) +
  geom_col(position = "stack", width = 0.8) +
  geom_text(data = assigned_labels,
            aes(label = sprintf("%.1f%%", pct), y = count / 2),
            size = 2.5, color = "white", fontface = "bold") +
  scale_fill_manual(values = fc_colors, drop = FALSE) +
  scale_y_continuous(labels = scales::label_comma()) +
  add_stand_facet() +
  labs(title = "featureCounts Read-pair Assignment",
       x = NULL, y = "Read pairs", fill = "Assignment status")

save_fig(p7, "fig7_featurecounts")
```

---

## Summary Table

```{r summary-table}
# Build per-sample summary row
fc_assigned <- fc_df %>%
  filter(!is.na(status)) %>%
  group_by(sample) %>%
  summarise(
    total_reads    = sum(count, na.rm = TRUE),
    assigned_reads = sum(count[status == "Assigned"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pct_assigned = assigned_reads / total_reads * 100)

summary_tbl <- sample_meta %>%
  select(sample, stand_type) %>%
  left_join(tiara_df %>% select(sample, n_total, n_euk),
            by = "sample") %>%
  left_join(metaeuk_df %>% select(sample, n_proteins), by = "sample") %>%
  left_join(heatmap_df %>% select(sample, pct_kofam, pct_cazy, pct_phi,
                                   pct_mmseqs, pct_fungi),
            by = "sample") %>%
  left_join(fc_assigned %>% select(sample, pct_assigned), by = "sample") %>%
  left_join(dbcan_df %>%
              filter(!is.na(n_hicof)) %>%
              group_by(sample) %>%
              summarise(n_cazy_hicof = sum(n_hicof), .groups = "drop"),
            by = "sample") %>%
  left_join(phi_df %>%
              filter(!is.na(n_hits)) %>%
              group_by(sample) %>%
              summarise(n_phi_hits = sum(n_hits), .groups = "drop"),
            by = "sample") %>%
  mutate(
    pct_euk = n_euk / n_total * 100
  ) %>%
  select(
    Sample    = sample,
    `Stand type` = stand_type,
    `Contigs (k)`  = n_total,
    `% euk`        = pct_euk,
    `Proteins`     = n_proteins,
    `% KO`         = pct_kofam,
    `CAZy (hi-cf)` = n_cazy_hicof,
    `PHI hits`     = n_phi_hits,
    `% MMseqs2 Fungi` = pct_fungi,
    `% reads assigned` = pct_assigned
  ) %>%
  mutate(`Contigs (k)` = round(`Contigs (k)` / 1000, 1))

kbl_tbl <- summary_tbl %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))

# Apply green→white gradient to % columns, blue gradient to count columns
color_cols <- c("% euk", "% KO", "% MMseqs2 Fungi", "% reads assigned",
                "Proteins", "CAZy (hi-cf)", "PHI hits")

tbl_out <- kable(kbl_tbl, caption = "Per-sample annotation metrics summary",
                 format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12)

for (cn in color_cols) {
  col_idx <- which(names(kbl_tbl) == cn)
  if (length(col_idx) == 1) {
    vals <- as.numeric(kbl_tbl[[cn]])
    if (any(!is.na(vals))) {
      tbl_out <- column_spec(tbl_out, col_idx,
                             background = spec_color(vals, na_color = "#F8F8F8",
                                                     option = "D", begin = 0.15, end = 0.85,
                                                     direction = 1))
    }
  }
}

tbl_out %>% scroll_box(width = "100%")
```

```{r save-summary-csv, include=FALSE}
write.csv(summary_tbl, file.path(out_dir, "qc_summary.csv"),
          row.names = FALSE, na = "")
```

---

*Report generated: `r Sys.time()`*
*Annotation directory: `r params$ann_dir`*
*Figures saved to: `r fig_dir`*
